Machine learning methods in streamflow prediction considering climate variable windowing

### 2021-06-28 paper/paper.tex
Figuras agora posicionadas apenas como htb
boxplot para o appendix, retirada do parágrafo sobre

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\linewidth]{Figures/mapa.pdf}
  \caption{Alto São Francisco sub-basin.}
  \label{fig:studyarea}
\end{figure}

@@

We collected atmospheric climate data generated by CPTEC/INPE, available on the Projeta Platform \cite{chou2014assessment,chou2014evaluation,Lyra2018}, to forecast streamflow for Três Marias reservoir. The available data includes the period from January 1999 to December 2005. We combined that with observed data of streamflow and water level for the same period, available on \cite{onsnivel,onsvazao}. In our study, streamflow is the target variable, but its previous value is also input to the prediction. All variables are shown in Table~\ref{tab:predictors}. For further information of data exploration, see Appendix A.

\begin{table}[htbp]
\centering
\caption{Relation of predictors.}
\label{tab:predictors}
    %\resizebox{0.6\columnwidth}{!}{%
    \renewcommand{\arraystretch}{.8}
    \small\begin{tabulary}{\textwidth}{lcCcCr}
        \toprule
            \normalsize\bfseries{Variable} &
            \normalsize\bfseries{Unit} &
            \normalsize\bfseries{Temporal resolution} &
            \normalsize\bfseries{Historical data}  \\
        \midrule
            Temperature (T)    & ºC   & Daily   & 1999-2005 \\
            Precipitation (P)  & mm   & Daily   & 1999-2005 \\
            Evaporation (E)    & mm   & Daily   & 1999-2005 \\
            Surface runoff (R) & mm   & Daily   & 1999-2005 \\
            Streamflow (SF)    & m³/s & Daily   & 1999-2005 \\
            Water level (WL)   & m    & Daily   & 1999-2005 \\
            Month (M)          & -    & -       & 1999-2005 \\
        \bottomrule
    \end{tabulary}
    %}
\end{table}

@@

\begin{figure}[htbp]
    \centering
    {\footnotesize\includesvg[width=.8\linewidth]{Diagrams/gridsearch.drawio}}
    \caption{Pipeline setup for machine learning model search}
    \label{fig:pipelines}
\end{figure}

@@

All variables, except month, were normalized when evaluating regressors that are sensitive to data scale, in order to prevent them from being dominated by variables with higher order of magnitude. We used the Min-Max normalization given by the Equation
    \begin{equation}
    \label{eqn:Normalization}
    x' = \frac{x-min(x)}{max(x)-min(x)}
    \end{equation}
to adjust the scale of the inputs. Where the value $x$ is the present value, $min(x)$ is the minimum value of the feature, and $max(x)$ is the maximum value of the feature. Then, we transformed that to their original scale.

While only the scale in the values changed with the normalization procedure, it is easier to identify the variation in mean and standard deviation between variables, as shown in Figure \ref{fig:Normalized} for the inputs in the first river.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth, trim=0cm 0 0 .7cm,clip=true]{Figures/Normalização.png}
  \caption{Boxplot of inputs from normalized data (left) and raw data (rifht)}
  \label{fig:Normalized}
\end{figure}

For the month variable, we applied one-hot encoding, which is often used for categorical variables, when the regressor was sensitive to the scale of the attributes, such as SVR, or Ordinal encoding, otherwise.

@@

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w01}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:r2_1}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w15}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:r2_15}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/R2_w30}
        \captionsetup{justification=justified,singlelinecheck=false}
        \captionsetup{margin={.23\linewidth, 0cm}}
        \caption{30 days windowing}
        \label{fig:r2_30}
    \end{subfigure}
    \caption{R-squared for \subref{fig:r2_1} 1, \subref{fig:r2_15} 15, and \subref{fig:r2_30} 30 days windowing}
    \label{fig:r2}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/MAE_w01}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:sfig1MAE}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/MAE_w15}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:sfig2MAE}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/MAE_w30}
        \captionsetup{margin={.23\linewidth, 0cm}}
        \captionsetup{justification=justified,singlelinecheck=false}
        \caption{30 days windowing}
        \label{fig:sfig3MAE}
    \end{subfigure}
    \caption{Mean Absolute Errors (MAE) for \subref{fig:sfig1MAE} 1, \subref{fig:sfig2MAE} 15, and \subref{fig:sfig3MAE} 30 days windowing}
    \label{fig:sfigMAE}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includesvg{Graphs/StackedComparisson}
    \caption{Comparison between true and predicted values for a prediction     horizon of 1 and 29 days. Visualization for a 2 year window, from 2000 to 2002}
    \label{fig:StackedComparisson}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/CategorizedImportance_w01}
        \captionsetup{margin={.17\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:CategorizedImportance01}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/CategorizedImportance_w15}
        \captionsetup{margin={.17\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:CategorizedImportance02}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/CategorizedImportance_w30}
        \captionsetup{margin={.26\linewidth, 0cm}}
        \captionsetup{justification=justified,singlelinecheck=false}
        \caption{30 days windowing}
        \label{fig:CategorizedImportance03}
    \end{subfigure}
    \caption{Variable importance for \subref{fig:CategorizedImportance01} 1, \subref{fig:CategorizedImportance02} 15, and \subref{fig:CategorizedImportance03} 30 days windowing}
    \label{fig:CategorizedImportance}
\end{figure}

>>>>

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.6\linewidth]{Figures/mapa.png}
  \includesvg{Figures/mapa_legend.drawio}
  \caption{Alto São Francisco sub-basin.}
  \label{fig:studyarea}
\end{figure}

@@

We collected atmospheric climate data generated by CPTEC/INPE, available on the Projeta Platform \cite{chou2014assessment,chou2014evaluation,Lyra2018}, to forecast streamflow for Três Marias reservoir. The available data includes the period from January 1999 to December 2005. We combined that with observed data of streamflow and water level for the same period, available on \cite{onsnivel,onsvazao}. In our study, streamflow is the target variable, but its previous value is also input to the prediction. All variables are shown in Table~\ref{tab:predictors}. For further information of data exploration, see Appendix~\ref{sec:DataExploration}.

\begin{table}[htb]
\centering
\caption{Relation of predictors.}
\label{tab:predictors}
    %\resizebox{0.6\columnwidth}{!}{%
    \renewcommand{\arraystretch}{.8}
    \small\begin{tabulary}{\textwidth}{lcCcCr}
        \toprule
            \normalsize\bfseries{Variable} &
            \normalsize\bfseries{Unit} &
            \normalsize\bfseries{Temporal resolution} &
            \normalsize\bfseries{Historical data}  \\
        \midrule
            Temperature (T)    & ºC   & Daily   & 1999-2005 \\
            Precipitation (P)  & mm   & Daily   & 1999-2005 \\
            Evaporation (E)    & mm   & Daily   & 1999-2005 \\
            Surface runoff (R) & mm   & Daily   & 1999-2005 \\
            Streamflow (SF)    & m³/s & Daily   & 1999-2005 \\
            Water level (WL)   & m    & Daily   & 1999-2005 \\
            Month (M)          & -    & -       & 1999-2005 \\
        \bottomrule
    \end{tabulary}
    %}
\end{table}

@@

\begin{figure}[htb]
    \centering
    {\footnotesize\includesvg[width=.8\linewidth]{Diagrams/gridsearch.drawio}}
    \caption{Pipeline setup for machine learning model search}
    \label{fig:pipelines}
\end{figure}

@@

All variables, except month, were normalized when evaluating regressors that are sensitive to data scale, in order to prevent them from being dominated by variables with higher order of magnitude. We used the Min-Max normalization given by the Equation
    \begin{equation}
    \label{eqn:Normalization}
    x' = \frac{x-min(x)}{max(x)-min(x)}
    \end{equation}
to adjust the scale of the inputs. Where the value $x$ is the present value, $min(x)$ is the minimum value of the feature, and $max(x)$ is the maximum value of the feature. Then, we transformed that to their original scale. For the month variable, we applied one-hot encoding, which is often used for categorical variables, when the regressor was sensitive to the scale of the attributes, such as SVR, or Ordinal encoding, otherwise.

@@

\begin{figure}[htb]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w01}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:r2_1}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w15}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:r2_15}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/R2_w30}
        \captionsetup{justification=justified,singlelinecheck=false}
        \captionsetup{margin={.23\linewidth, 0cm}}
        \caption{30 days windowing}
        \label{fig:r2_30}
    \end{subfigure}
    \caption{R-squared for \subref{fig:r2_1} 1, \subref{fig:r2_15} 15, and \subref{fig:r2_30} 30 days windowing}
    \label{fig:r2}
\end{figure}

@@

\begin{figure}[htb]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/MAE_w01}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:sfig1MAE}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/MAE_w15}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:sfig2MAE}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/MAE_w30}
        \captionsetup{margin={.23\linewidth, 0cm}}
        \captionsetup{justification=justified,singlelinecheck=false}
        \caption{30 days windowing}
        \label{fig:sfig3MAE}
    \end{subfigure}
    \caption{Mean Absolute Errors (MAE) for \subref{fig:sfig1MAE} 1, \subref{fig:sfig2MAE} 15, and \subref{fig:sfig3MAE} 30 days windowing}
    \label{fig:sfigMAE}
\end{figure}

@@

\begin{figure}[htb]
    \centering
    \includesvg{Graphs/StackedComparisson}
    \caption{Comparison between true and predicted values for a prediction     horizon of 1 and 29 days. Visualization for a 2 year window, from 2000 to 2002}
    \label{fig:StackedComparisson}
\end{figure}

@@

\begin{figure}[htb]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/CategorizedImportance_w01}
        \captionsetup{margin={.17\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:CategorizedImportance01}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/CategorizedImportance_w15}
        \captionsetup{margin={.17\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:CategorizedImportance02}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/CategorizedImportance_w30}
        \captionsetup{margin={.26\linewidth, 0cm}}
        \captionsetup{justification=justified,singlelinecheck=false}
        \caption{30 days windowing}
        \label{fig:CategorizedImportance03}
    \end{subfigure}
    \caption{Variable importance for \subref{fig:CategorizedImportance01} 1, \subref{fig:CategorizedImportance02} 15, and \subref{fig:CategorizedImportance03} 30 days windowing}
    \label{fig:CategorizedImportance}
\end{figure}
